---
title: 'Review: Matrix decomposition'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 2.5, fig.height = 2)
```

### Helper functions for visualization

```{r, warning=FALSE}
library(ggplot2)
heatmap = function(A){
  n = nrow(A)
  p = ncol(A)  
  df = data.frame(value = c(A),  i = 1:n, j = rep(1:p, rep(n, p)))
  ggplot(df, aes(j, i, fill = value)) + 
    geom_tile(color = "black")+
    scale_fill_gradient2(low="#619CFF", mid="white", high="#F8766D")+
    scale_y_reverse()+
    theme_void()
}

barplot = function(values){
  n = length(values)
  df = data.frame(value = values,  index = 1:n)
  ggplot(df, aes(index, value, fill = value)) + 
    geom_bar(color = "black", stat = "identity")+
    scale_fill_gradient2(low="#619CFF", mid="white", high="#F8766D")+
    theme_bw()
}
```

\newpage

### Generating matrices with various structures

Generate a random matrix.
```{r}
n = 5
p = 3
A = matrix(rnorm(n * p), n, p)
A
```


\newpage

Find QR decomposition. QR can be used to generate a random column orthogonal matrix.
```{r}
QR = qr(A)
Q = qr.Q(QR)
Q
t(Q) %*% Q
heatmap(t(Q) %*% Q)
heatmap(Q %*% t(Q))
R = qr.R(QR)
R
heatmap(R)
```

\newpage 

Generate a symmetric matrix.
```{r}
B = matrix(rnorm(n * n), n, n)
B = t(B) + B
heatmap(B)
```

\newpage

Compute eigen decomposition.
```{r}
ED = eigen(B)
lambdas = ED$values
lambdas
barplot(lambdas)
U = ED$vectors
heatmap(t(U) %*% U)
heatmap(U %*% t(U))
```

\newpage 

Generate a rank-$r$ matrix.
```{r}
r = 1
A = matrix(rnorm(n * r), n, r)
B = matrix(rnorm(p * r), p, r)
C = A %*% t(B)
QR = qr(C)
QR$rank
heatmap(qr.R(QR))
```

\newpage

Generate a PSD matrix.
```{r}
A = matrix(rnorm(n * p), n, p)
C = A %*% t(A)
heatmap(C)
barplot(eigen(C)$values)
```

\newpage

Generate a PD matrix.
```{r}
A = matrix(rnorm(n * n), n, n)
C = t(A) %*% A
eigen(C)$values
barplot(eigen(C)$values)
```

\newpage

Find square root of a matrix.
```{r}
ED = eigen(C)
Csqrt = ED$vectors %*% diag(sqrt(ED$values)) %*% t(ED$vectors)
heatmap(Csqrt %*% Csqrt)
heatmap(C)
```

\newpage

Generate a matrix with given eigenvalues.
```{r}
A = matrix(rnorm(n * n), n, n)
U = qr.Q(qr(A))
lambdas = c(1, 2, 3, 4, 4)
C = U %*% diag(lambdas) %*% t(U)
barplot(eigen(C)$values)
```

\newpage

Generate a projection matrix.
```{r}
A = matrix(rnorm(n * p), n, p)
P = A %*% solve(t(A) %*% A) %*% t(A)
heatmap(P)
barplot(eigen(P)$values)
barplot(diag(P))
```

\newpage

Find orthogonal complement to a basis
```{r}
A = matrix(rnorm(n * p), n, p)
Q = qr.Q(qr(A))
P = Q %*% t(Q)
barplot(eigen(P)$values)
Pperp = diag(n) - P
barplot(eigen(Pperp)$values)
Qperp = eigen(Pperp)$vector[, 1:(n - p)]
t(Qperp) %*% Q
```

\newpage

Compute singular value decomposition.
```{r}
A = matrix(rnorm(n * p), n, p)
SVD = svd(A)
U = SVD$u
V = SVD$v
d = SVD$d
heatmap(U)
heatmap(V)
barplot(d)
```

\newpage

What if p>n?
```{r}
A = t(A)
SVD = svd(A)
U = SVD$u
V = SVD$v
d = SVD$d
heatmap(U)
heatmap(V)
barplot(d)
```

\newpage

Link between SVD and ED.
```{r}
A = matrix(rnorm(n * p), n, p)
svd(A)$v
eigen(t(A) %*% A)$vectors
svd(A)$u
eigen(A %*% t(A))$vectors
```

\newpage

### Centering and computing sample covariance matrix

Matrix view for centering.
```{r}
X = matrix(rnorm(n * p), n, p)
xbar = colMeans(X)
xbar
Xc = scale(X, center = T, scale = F)
colMeans(Xc)
C = diag(n) - matrix(1/n, n, n)
heatmap(C)
Xc = C %*% X
colMeans(Xc)
```


\newpage

Sample covariance matrix.
```{r}
S = (t(X) %*% C %*% X)/(n - 1)
heatmap(S)
heatmap(cov(X))
```


