---
title: 'Low-rank matrix apploximation'
output:
  html_document:
    df_print: kable
  pdf_document: default
---

```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 4.5, warning = F, message = F)
theme_set(theme_bw())
```

### Helper functions

```{r, warning=FALSE}
library(ggplot2)
library(dplyr)
library(viridis)
library(ggfortify)

barplot = function(values){
  n = length(values)
  df = data.frame(value = values,  index = 1:n)
  ggplot(df, aes(index, value, fill = value)) + 
    geom_bar(color = "black", stat = "identity")+
    scale_fill_gradient2(low="#619CFF", mid="white", high="#F8766D")+
    theme_bw()+
    theme(legend.position = "none")
}

heatmap = function(A){
  n = nrow(A)
  p = ncol(A)  
  df = data.frame(value = c(A),  i = 1:n, j = rep(1:p, rep(n, p)))
  ggplot(df, aes(j, i, fill = value)) + 
    geom_tile()+
    scale_fill_viridis()+
    #scale_fill_gradient(low="grey", high="black")+
    scale_y_reverse()+
    theme_void()
}

Fnorm = function(A) return(sqrt(sum(A^2)))
```


\newpage

### Visualizing low-rank matrices

```{r}
n = 50
p = 30
r = 1
A = matrix(runif(n*r), n, r)
B = matrix(runif(p*r), p, r)
X = A %*% t(B)
heatmap(X)
```

```{r}
r = 2
A = matrix(runif(n*r), n, r)
B = matrix(runif(p*r), p, r)
X = A %*% t(B)
heatmap(X)
```
```{r}
r = 10
A = matrix(runif(n*r), n, r)
B = matrix(runif(p*r), p, r)
X = A %*% t(B)
heatmap(X)
```
```{r}
r = 30
A = matrix(runif(n*r), n, r)
B = matrix(runif(p*r), p, r)
X = A %*% t(B)
heatmap(X)
```


### Low-rank approximation

```{r}
library(tiff)
library(rasterImage)
LRA = function(X, r){
  SVD = svd(X)
  Xhat = SVD$u[,1:r] %*% diag(SVD$d[1:r],r) %*% t(SVD$v[,1:r])
  return(Xhat)
}
```

### Image compression

Load and image and check how rank impacts the resulting image.
```{r}
X = readTIFF("img/baboon.tiff")[,,1]
X = scale(X, center = T, scale = F)
dim(X)
ve = (svd(X)$d)^2/sum((svd(X)$d)^2)
barplot(cumsum(ve)[1:50])+
  xlab("principal component")+
  ylab("cumulative variance explained")
heatmap(X)
heatmap(LRA(X, 1))
heatmap(LRA(X, 5))
heatmap(LRA(X, 10))
heatmap(LRA(X, 50))
```

### De-noising of an image

Create a low-rank pattern. 
```{r}
set.seed(3)
library(splines)
n = 100
df = 15
H = ns(1:n, df = df)
Theta = matrix(rnorm(df * 3), df, 3)
A = H %*% Theta
X = A %*% t(A)
heatmap(X)
ve = (svd(X)$d)^2/sum((svd(X)$d)^2)
barplot(ve[1:10])+
  xlab("principal component")+
  ylab("variance explained")
```

Add noise and try de-noising varying the rank of approximation.
```{r}
E = matrix(rnorm(n*n, 0, 0.2), n, n)
Xnoise = X+E
heatmap(Xnoise)
ve = (svd(Xnoise)$d)^2/sum((svd(Xnoise)$d)^2)
barplot(ve[1:10])+
  xlab("principal component")+
  ylab("variance explained")
heatmap(LRA(Xnoise, 1))
heatmap(LRA(Xnoise, 3))
heatmap(LRA(Xnoise, 20))
```

Increase the noise and check the performance.
```{r}
E = matrix(rnorm(n*n, 0, 0.5), n, n)
Xnoise = X+E
heatmap(Xnoise)
ve = (svd(Xnoise)$d)^2/sum((svd(Xnoise)$d)^2)
barplot(ve[1:10])+
  xlab("principal component")+
  ylab("variance explained")
heatmap(LRA(Xnoise, 1))
heatmap(LRA(Xnoise, 3))
heatmap(LRA(Xnoise, 20))
```

### Soft rank approximation

```{r}
nucnorm = function(X){
  d = svd(X)$d
  return(sum(d))
}
SRA = function(X, lambda){
  SVD = svd(X)
  d = SVD$d
  dstar = pmax(d - lambda, 0)
  return(SVD$u %*% diag(dstar) %*% t(SVD$v))
}
```


Let's return to baboons.
```{r}
X = readTIFF("img/baboon.tiff")[,,1]
X = scale(X, center = T, scale = F)
heatmap(X)
n = nrow(X)
p = ncol(X)
```


Check the distribution of singular values after applying the soft-threshold.
```{r, fig.width = 6}
singvals = c()
for(lambda in c(0.1, 1, 2, 5, 10)){
  Xhat = SRA(X, lambda)
  singvals = rbind(singvals, data.frame(lambda, dstar = svd(Xhat)$d, index = 1:p))
}
ggplot(singvals %>% mutate(lambda = as.factor(lambda)))+
  geom_bar(aes(x = index, y = dstar, fill = lambda), stat = "identity",  alpha = 0.5)+
  facet_grid(lambda~., labeller = label_both)+
  theme(legend.position = "none")+
  ylab("signular values of the approximation")
```


Check that the nuclear norm decreases.
```{r}
singvals %>% 
  group_by(lambda) %>%
  summarise(norm = nucnorm(dstar)) %>%
ggplot(aes(lambda, norm))+
  geom_point()+
  geom_line()+
  ylab("nuclear norm")+
  xlab(expression(lambda))
```

The results resemble the ones obtained by low-rank matrix approximation.

```{r}
heatmap(SRA(X, 1))
heatmap(SRA(X, 5))
heatmap(SRA(X, 10))
```






