---
title: 'PCA variations'
output:
  html_document:
    df_print: kable
  pdf_document: default
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 4.5, warning = F, message = F)
theme_set(theme_bw())
```

### Helper functions

```{r, warning=FALSE}
library(ggplot2)
library(dplyr)
library(viridis)
library(ggfortify)
library(tictoc)

barplot = function(values){
  n = length(values)
  df = data.frame(value = values,  index = 1:n)
  ggplot(df, aes(index, value)) + 
    geom_bar(stat = "identity", color = "orange")+
    theme_bw()+
    theme(legend.position = "none")
}

heatmap = function(A){
  n = nrow(A)
  p = ncol(A)  
  df = data.frame(value = c(A),  i = 1:n, j = rep(1:p, rep(n, p)))
  ggplot(df, aes(j, i, fill = value)) + 
    geom_tile()+
    scale_fill_viridis()+
    #scale_fill_gradient(low="grey", high="black")+
    scale_y_reverse()+
    theme_void()
}

Fnorm = function(A) return(sqrt(sum(A^2)))
```


### PCA with missing values

```{r, include = F}
heatmap = function(A){
  n = nrow(A)
  p = ncol(A)  
  df = data.frame(value = c(A),  i = 1:n, j = rep(1:p, rep(n, p)))
  ggplot(df, aes(j, i, fill = value)) + 
    geom_tile()+
    scale_fill_gradient(low="green", high = "red")+
    scale_y_reverse()+
    theme_void()
}
```

Load microarray data.
```{r}
microarray = read.csv("data/cancer_microarray.csv")
X = microarray %>% select(-cancer_type) %>% as.matrix()
label = microarray$cancer_type
dim(X)
heatmap(X)
```

Remove columns with missing values and check the PCA results.
```{r}

Xremove = t(na.omit(t(X)))
dim(Xremove)
PCA =  prcomp(Xremove, scale = F, center = T)
autoplot(PCA, data = microarray, shape = F, color = "cancer_type", label = TRUE, label.label = "cancer_type", scale = 0)+
  theme(legend.position = "none")
```
```{r, include = F}
heatmap = function(A){
  n = nrow(A)
  p = ncol(A)  
  df = data.frame(value = c(A),  i = 1:n, j = rep(1:p, rep(n, p)))
  ggplot(df, aes(j, i, fill = value)) + 
    geom_tile()+
    scale_fill_gradient(low="white", high = "black")+
    scale_y_reverse()+
    theme_void()
}
```

Hard-impute step.
```{r}
W = 1 - is.na(X)
heatmap(W)
hardimpute_iter = function(X, r, Xhat){
  X[is.na(X)] = 0
  Y = W * X + (1 - W) * Xhat
  SVD = svd(Y)
  Xhat = SVD$u[,1:r] %*% diag(SVD$d[1:r], r) %*% t(SVD$v[,1:r])
  return(Xhat)
}
```

Run hard-impute for 20 iterations.
```{r}
hardimpute = function(X, r, verbose = T){
  n = nrow(X)
  p = ncol(X)
  Xhat = matrix(rnorm(n*p), n, p)
  for(iter in 1:20){
    Xhat0 = Xhat
    Xhat = hardimpute_iter(X, r, Xhat)
    rch = sum((Xhat - Xhat0)^2)/sum((Xhat0)^2)
    if(verbose) cat("\niter:", iter, "| relative change in X:", rch)
  }
  return(Xhat)
}
```

Use $r=2$ to impute missing values.
```{r}
Ximp = hardimpute(X, 2)
```

Check the PCA results.
```{r}
PCA =  prcomp(Ximp, scale = T, center = T)
autoplot(PCA, data = microarray, shape = F, color = "cancer_type", label = TRUE, label.label = "cancer_type", scale = 0)+
  theme(legend.position = "none")
```

Try other values for $r$.
```{r}
performance = c()
for(r in seq(2, 50, 2)){
  Ximp = hardimpute(X, r, verbose = F)
  relerr = sum((X - Ximp)^2, na.rm = T)/sum(X^2, na.rm = T)
  performance = rbind(performance, data.frame(r, relerr))
}
ggplot(performance, aes(r, relerr))+
  geom_point()+
  geom_line()+
  xlab("rank")+
  ylab("relative approximation error")
  
```

### PCA with sparsity

Load data.
```{r}
microarray = read.csv("data/cancer_microarray_full.csv") %>% mutate(cancer_type = as.factor(cancer_type))
X = microarray %>% select(-cancer_type) %>% as.matrix()
label = microarray$cancer_type
n = nrow(X)
p = ncol(X)
dim(X)
```

Remove columns with missing values and check the PCA results.
```{r}
Xc = scale(X, center = T, scale = F)
PCA =  prcomp(Xc, scale = F, center = T)
autoplot(PCA, data = microarray, shape = F, color = "cancer_type", label = TRUE, label.label = "cancer_type", scale = 0)+
  theme(legend.position = "none")
```


The first PC loading vector would look like this.

```{r, fig.width = 10}
v1 = PCA$rotation[,1]
barplot(v1)+
  xlab("gene")+
  ylab("loading value")+
  geom_hline(aes(yintercept = c(0.05)), linetype = "dashed", color = "black")+
  geom_hline(aes(yintercept = c(-0.05)), linetype = "dashed", color = "black")
```

```{r, include = F}
heatmap = function(A){
  n = nrow(A)
  p = ncol(A)  
  df = data.frame(value = c(A),  i = 1:n, j = rep(1:p, rep(n, p)))
  ggplot(df, aes(j, i, fill = value)) + 
    geom_tile()+
    scale_fill_gradient(low="green", high = "red")+
    scale_y_reverse()+
    theme_void()
}
```


Visualize the "important" genes.
```{r}
pick = (abs(v1) > 0.05)
dim(X[,pick])
heatmap(X[,pick])
```
Run PCA on important genes only.
```{r}
PCA =  prcomp(X[,pick], scale = F, center = T)
autoplot(PCA, data = microarray, shape = F, color = "cancer_type", label = TRUE, label.label = "cancer_type", scale = 0)+
  theme(legend.position = "none")
```

Use `PMA` package to find sparse PCA solution. Try different thresholds for the $\|v\|_1$.
```{r}
library(PMA)
sqrt(p)
spc3 = SPC(Xc, sumabsv = 3, K = 2, orth = T, niter = 100)
spc3
heatmap(X[,spc3$v[,1] != 0])
spc7 = SPC(Xc, sumabsv = 7, K = 2, orth = T, niter = 50)
spc7
heatmap(X[,spc7$v[,1] != 0])
spc12 = SPC(Xc, sumabsv = 12, K = 2, orth = T, niter = 50)
spc12
heatmap(X[,spc12$v[,1] != 0])
```


Plot the loading vectors for the first PC.
```{r, fig.width=10}
df = data.frame(loading = c(spc3$v[,1], spc7$v[,1], spc12$v[,1]), threshold = rep(c(3, 7, 12), rep(ncol(X), 3)), gene = 1:p)
ggplot(df, aes(gene, loading)) + 
  geom_bar(stat = "identity", color = "orange")+
  theme_bw()+
  theme(legend.position = "none")+
  facet_grid(threshold~., labeller = label_both)
```

Plot the biplots.
```{r, fig.width=10}
Z = rbind(Xc %*% spc3$v, Xc %*% spc7$v, Xc %*% spc12$v)
colnames(Z) = c("PC1", "PC2")
df = data.frame(Z, threshold = rep(c(3, 5, 12), rep(nrow(X), 3)), sample = 1:n, cancer_type = label)
ggplot(df, aes(PC1, PC2, color = cancer_type, label = cancer_type)) + 
  geom_text()+
  stat_ellipse()+
  theme_bw()+
  theme(legend.position = "none")+
  facet_wrap(~threshold, scale = "free", ncol = , labeller = label_both)
```

### Kernel PCA

#### Two concentric circles

Generate data.
```{r}
n = 100
r1 = 0.3
r2 = 1
X1 = matrix(rnorm(n * 2), n, 2)
X1 = X1/sqrt(rowSums(X1^2)) * r1
X2 = matrix(rnorm(n * 2), n, 2)
X2 = X2/sqrt(rowSums(X2^2)) * r2
E = matrix(rnorm(2*n * 2, 0, 0.05), 2*n, 2)
X = rbind(X1, X2) + E
X = scale (X, scale = F, center = T)
colnames(X) = c("x1", "x2")
labels = c(rep(1, n), rep(2, n))
df = data.frame(X, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")
```

Apply PCA.
```{r}
PCA = prcomp(X)
Xhat = PCA$x
colnames(Xhat) = c("x1", "x2")
df = data.frame(Xhat, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  xlab("PC1")+
  ylab("PC2")
```

Apply kernel PCA with the RBF kernel.
```{r}
library(kernlab)
KPCA = kpca(X, kernel="rbfdot", kpar=list(sigma = 10), features=2)
Xhat = rotated(KPCA)
colnames(Xhat) = c("x1", "x2")
df = data.frame(Xhat, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  xlab("PC1")+
  ylab("PC2")
```


#### Two half-circles

Generate data.
```{r}
n = 100
theta = -pi / 180 * 10
R = matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2)
X1 = matrix(rnorm(n * 2), n, 2)
X1 = X1/sqrt(rowSums(X1^2))
X1[,2] = abs(X1[,2])
X1 = scale(X1 %*% R, center = c(-0.6, 0), scale = F)
X2 = matrix(rnorm(n * 2), n, 2)
X2 = X2/sqrt(rowSums(X2^2))
X2[,2] = -abs(X2[,2])
X2 = scale(X2 %*% R, center = c(0.6, 0), scale = F)
E = matrix(rnorm(2*n * 2, 0, 0.03), 2*n, 2)
X = rbind(X1, X2) + E
X = scale(X, scale = F, center = T)
colnames(X) = c("x1", "x2")
labels = c(rep(1, n), rep(2, n))
df = data.frame(X, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```

Apply PCA.
```{r}
PCA = prcomp(X)
Xhat = PCA$x
colnames(Xhat) = c("x1", "x2")
df = data.frame(Xhat, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  xlab("PC1")+
  ylab("PC2")
```

Apply kernel PCA with the RBF kernel.
```{r}
KPCA = kpca(X, kernel="rbfdot", kpar=list(sigma = 7), features=2)
Xhat = rotated(KPCA)
colnames(Xhat) = c("x1", "x2")
df = data.frame(Xhat, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  xlab("PC1")+
  ylab("PC2")
```


### tSNE


#### Distances between clusters and cluster sizes are not informative.

Generate data: three clusters of different size.
```{r}
n = 100
X1 = matrix(rnorm(n * 2, 0, 0.05), n, 2)
X2 = matrix(rnorm(n * 2, 0, 0.2), n, 2)
X2[,1] = X2[,1] + 1
X3 = matrix(rnorm(n * 2, 0, 0.3), n, 2)
X3[,1] = X3[,1] + 3
X = rbind(X1, X2, X3)
colnames(X) = c("x1", "x2")
labels = c(rep(1, n), rep(2, n), rep(3, n))
df = data.frame(X, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  ylim(-3,3)+
  coord_fixed()
```

Try different values of `perplexity`.
```{r}
library(tsne)
TSNE = tsne(X, perplexity = 5)
colnames(TSNE) = c("x1", "x2")
df = data.frame(TSNE, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```
  
```{r}
library(tsne)
TSNE = tsne(X, perplexity = 10)
colnames(TSNE) = c("x1", "x2")
df = data.frame(TSNE, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```

```{r}
library(tsne)
TSNE = tsne(X, perplexity = 50)
colnames(TSNE) = c("x1", "x2")
df = data.frame(TSNE, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```

#### Two concentric circles


Generate data.
```{r}
n = 500
r1 = 0.3
r2 = 1
X1 = matrix(rnorm(n * 2), n, 2)
X1 = X1/sqrt(rowSums(X1^2)) * r1
X2 = matrix(rnorm(n * 2), n, 2)
X2 = X2/sqrt(rowSums(X2^2)) * r2
E = matrix(rnorm(2*n * 2, 0, 0.1), 2*n, 2)
X = rbind(X1, X2) + E
X = scale (X, scale = F, center = T)
colnames(X) = c("x1", "x2")
labels = c(rep(1, n), rep(2, n))
df = data.frame(X, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")
```


Let's try UMAP.
```{r}
library(umap)
UMAP = umap(X)
colnames(UMAP$layout) = c("x1", "x2")
df = data.frame(UMAP$layout, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```

Change the parameters. First, change `n_neighbor`. 
```{r}
custom.config = umap.defaults
custom.config
custom.config$n_neighbors = 5
UMAP = umap(X, config = custom.config)
colnames(UMAP$layout) = c("x1", "x2")
df = data.frame(UMAP$layout, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```

```{r}
custom.config$n_neighbors = 50
UMAP = umap(X, config = custom.config)
colnames(UMAP$layout) = c("x1", "x2")
df = data.frame(UMAP$layout, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```

Then change `min_dist`.
```{r}
custom.config = umap.defaults
custom.config$min_dist = 0.9
UMAP = umap(X, config = custom.config)
colnames(UMAP$layout) = c("x1", "x2")
df = data.frame(UMAP$layout, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```
```{r}
custom.config$min_dist = 0.01
UMAP = umap(X, config = custom.config)
colnames(UMAP$layout) = c("x1", "x2")
df = data.frame(UMAP$layout, label = as.factor(labels))
ggplot(df, aes(x1, x2, color = label))+
  geom_point()+
  theme(legend.position = "none")+
  coord_fixed()
```


#### tSNE and UMAP for digits

We first check how many digits are in the training data.
```{r}
zip = read.table("Data/zip.train")
colnames(zip) = c("label", paste0("pixel", 1:256))
zip = zip %>% mutate(label = factor(label))
digits = zip[,-1]
```


Plot for the first two principal components. 
```{r, fig.height = 8, fig.width = 8}
PCA = prcomp(digits)
autoplot(PCA, data = zip, shape = F, color = "label", label = TRUE, label.label = "label", scale = 0)+
  theme(legend.position = "none")
```



Let's try tSNE and plot the result.
```{r, fig.height = 8, fig.width = 8}
tic()
TSNE = tsne(digits, perplexity = 30)
tic()
colnames(TSNE) = c("x1", "x2")
df = data.frame(TSNE, label = as.factor(zip$label))
ggplot(df, aes(x1, x2, color = label, label = label))+
  geom_text()+
  theme(legend.position = "none")+
  coord_fixed()
```

Now, we try UMAP. 
```{r, fig.height = 8, fig.width = 8}
tic()
UMAP = umap(digits)
toc()
colnames(UMAP$layout) = c("x1", "x2")
df = data.frame(UMAP$layout, label = as.factor(zip$label))
ggplot(df, aes(x1, x2, color = label, label = label))+
  geom_text()+
  theme(legend.position = "none")+
  coord_fixed()
```

