---
title: 'Clustering: K-means'
output:
  html_document:
    df_print: kable
  pdf_document: default
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 4.5, warning = F, message = F)
theme_set(theme_bw())
```

### Helper functions

```{r, warning=FALSE}
library(ggplot2)
library(dplyr)
library(viridis)
library(ggfortify)
library(tictoc)

scatterplot = function(X, M, cluster, label = F){
  if(length(unique(cluster)) == 1){
    plt = ggplot()+
      geom_point(X, mapping = aes(longtitude, latitude))+
      geom_point(M, mapping = aes(longtitude, latitude), shape = 4, size = 4)
    if(label) return(plt + geom_text(X, mapping = aes(longtitude, latitude, label = 1:nrow(X)), nudge_x = 0.1))
    else return(plt)
  }
  else{
    ggplot()+
      geom_point(data.frame(X, cluster = as.factor(cluster)), mapping = aes(longtitude, latitude, color = cluster))+
      geom_point(M, mapping = aes(longtitude, latitude), shape = 4, size = 4)+
      theme(legend.position = "none")
  } 
}

barplot = function(values){
  n = length(values)
  df = data.frame(value = values,  index = 1:n)
  ggplot(df, aes(index, value, fill = value)) + 
    geom_bar(color = "black", stat = "identity")+
    scale_fill_gradient2(low="#619CFF", mid="white", high="#F8766D")+
    scale_x_continuous(breaks = 1:n)+
    theme_bw()+
    theme(legend.position = "none")
}
```


### Urban planning

Imagine we have a location with $n = 18$ houses and the city administration wants to build $K = 3$ grocery shops. 
```{r}
set.seed(6)
n = 18
K = 3
M = matrix(rnorm(K * 2), K, 2)
colnames(M) =  c("longtitude", "latitude")
Ms = M[rep(c(1,2,3), times = c(n/6, n/3, n/2)),]
X = matrix(rnorm(n * 2, 0, 0.3), n, 2) + Ms
colnames(X) = c("longtitude", "latitude")
plt = ggplot()+
  geom_point(X, mapping = aes(longtitude, latitude))
plt
```



We will use K-means with $K = 3$ to identify the locations for the grocery stores. 
```{r}
KM = kmeans(X, centers = K, algorithm = "Lloyd")
KM$cluster
KM$centers
KM$size
KM$iter
```


Let's investigate the impact of initialization on the results.
```{r}
M0 = M - 0.6
scatterplot(X, M0, rep(1, n))
for(iter in 1:3){
  KM = kmeans(X, iter.max = iter, centers = M0, algorithm = "Lloyd")
  print(scatterplot(X, KM$centers, KM$cluster))
}
```

Now we try different initialization. The result is quite different.
```{r}
M0 = M - 1
scatterplot(X, M0, rep(1, n))
for(iter in 1:3){
  KM = kmeans(X, iter.max = iter, centers = M0, algorithm = "Lloyd")
  print(scatterplot(X, KM$centers, KM$cluster))
}
```

### K-means++

The first centroid $m_1$ is chosen to be $x_i$ at random.
```{r}
set.seed(0)
M1 = X[sample(1:n, 1),,drop = F]
scatterplot(X, M1, rep(1, n), label = T)
```

The next centroid is chosen to be $x_i$ with probability proportional to $\|x_i-m_{k-1}\|^2$.
```{r}
D = as.matrix(dist(rbind(M1, X)))[2:(n+1),1]
P = D^2/sum(D^2)
barplot(P)+
  ylab("probability")
pick = sample(1:n, 1, prob = P)
M2 = X[pick,,drop = F]
scatterplot(X, rbind(M1, M2), rep(1, n), label = T)
D = as.matrix(dist(rbind(M2, X)))[2:(n+1),1]
P = D^2/sum(D^2)
barplot(P)+
  ylab("probability")
pick = sample(1:n, 1, prob = P)
M3 = X[pick,,drop = F]
scatterplot(X, rbind(M1, M2, M3), rep(1, n), label = T)
```


Run K-means with the K-means++ initialization.
```{r}
M0 = rbind(M1, M2, M3)
KM = kmeans(X, centers = M0, algorithm = "Lloyd")
print(scatterplot(X, KM$centers, KM$cluster))
```

### Between- and within-cluster scatter


Check the pattern for $B(K)$ and $W(K)$.
```{r}
Ks = 2:10
Bs = c()
Ws = c()
for(K in Ks){
  KM = kmeans(X, centers = K, iter.max = 10, nstart = 10)
  Bs = c(Bs, KM$betweenss)
  Ws = c(Ws, KM$tot.withinss)
}
df = data.frame(scatter = c(Bs, Ws, Bs + Ws), K = Ks, type = rep(c("between", "within", "total"), times = rep(length(Ks), 3)))
ggplot(df, aes(K, scatter, color = type))+
  geom_point()+
  geom_line()
```


# Gap statistics

Compute the Gap statistics.
```{r}
library(factoextra)
library(cluster)
gapstat = clusGap(X, FUN = kmeans, nstart = 50, K.max = 10, B = 50)
print(gapstat, method = "Tibs2001SEmax")
fviz_gap_stat(gapstat, maxSE = list(method = "Tibs2001SEmax", SE.factor = 1))
```

Make clusters more separated from each other.
```{r}
set.seed(6)
n = 18
K = 3
M = matrix(rnorm(K * 2), K, 2)
colnames(M) =  c("longtitude", "latitude")
Ms = M[rep(c(1,2,3), times = c(n/6, n/3, n/2)),]
X = matrix(rnorm(n * 2, 0, 0.1), n, 2) + Ms
colnames(X) = c("longtitude", "latitude")
ggplot()+
  geom_point(X, mapping = aes(longtitude, latitude))
gapstat = clusGap(X, FUN = kmeans, nstart = 50, K.max = 10, B = 50)
fviz_gap_stat(gapstat, maxSE = list(method = "Tibs2001SEmax", SE.factor = 1))
```

